\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
Updated manuscript for GLM and GLMM notation

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
\label{eq:linear-model}
\end{equation}

@@

\begin{equation}
\eta_{ijt}=\beta_0+\beta_{j}x_{ijt},
\label{eq:GLM-eta}
\end{equation}

@@

\begin{equation}
\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\alpha_{ij},
\label{eq:GLMM-eta}
\end{equation}

@@

Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group, and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\). A GAM version of a linear model can be written as:

\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
  \label{eq:GAM}
\end{equation}

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for \(f(x_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(x_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

@@

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the second derivative of the spline. The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} To better understand the smooth fitting process, we can use a physics analogy.

Suppose that we are traveling by car through a mountainous country and that we want to map our route (using a GPS device) on a specific road that only has one lane and that goes through a mountain range within that country; moreover, we want to see how our route changes if we want to optimize fuel consumption. In this analogy, the exact middle of the road is ``the truth'' or the ``mean function'' which we want to approximate through the route we will create. Moreover, because we know the ``truth'' in advance, we know that the road has many turns and ascends and descends multiple times, but overall it does it on a ``smooth'' way. The first time we travel the road, we impose no restrictions on fuel consumption (similar to setting \(\lambda \rightarrow 0\)), and we are constantly driving on the road shoulder, making abrupt stops without apparent reason, and also accelerating the vehicle suddenly at irregular intervals. When the trip is completed and the route taken is displayed along the ``truth'' on the GPS device, we notice that our route was ``rough'' and with a lot of ``noise'' because the trajectory of the vehicle deviated from the ``truth'' many times.

Besides being a certainly unpleasant driving experience, we know that such route does not accurately approximate the known ``truth.'' Because of this, we drive on the same road again, but this time to achieve our goal of a smooth drive, we want to optimize our fuel consumption by penalizing the acceleration of the vehicle (the second derivative of position with respect to time). Practically, this translates to maintaining constant speed as much as possible (which would also penalize sudden stops and acceleration bursts) while also staying as much as possible in the center of lane. We repeat the trip again and decide to set a mean value for our speed, which results in a certain fuel consumption value. If we repeat this process many times, we will eventually find a value of mean speed that produces the best overall fuel consumption (we find an optimal value for \(\lambda\)). This in a crude sense, is the process that we use to penalize the wiggliness of the smooth.

With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A. The four basis functions (and the intercept) are shown. Each of the basis functions is evaluated at six different points (because there are six points on the timeline). To control the ``wiggliness'' of the fit, the coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\), which is multiplied by the integrated square of the second derivative of the spline expansion. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate. The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.

@@

Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process. Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie. In a ``point-wise'' CI for non-linear data, the ``true'' function will lie outside of the CI in certain regions at a given frequency (because the data does not follow a linear trend). This means that if a point-wise CI is obtained for 100 random samples, the total number of CIs that contain the true function \emph{entirely} is much less than the nominal value of the CI (95\%, usually).

Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data). For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

>>>>

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
\label{eq:linear-model}
\end{equation}

@@

\begin{equation}
\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j},
\label{eq:GLM-eta}
\end{equation}

@@

\begin{equation}
\eta_{ijt}=\beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij},
\label{eq:GLMM-eta}
\end{equation}

@@

Where \(\beta_{0}\) is the intercept, \(\beta_{1}\) is the coefficient for each treatment group. Notice that we and \(f(time_{t}|\beta_{j})\) represents the smooth function of time, with coefficient \(\beta_{j}\). A GAM version of a linear model can be written as:

\begin{equation}
  y_{ijt}=\beta_0+f(time_t\mid \beta_j)+\varepsilon_{ijt}
  \label{eq:GAM}
\end{equation}

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and the change of \(y_{ijt}\) over time (instead of being defined by \(\beta_{j}time_{ijt}\) as in Equation \eqref{eq:GLM-eta}) is represented by the smooth function \(f(time_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the deviation of each observation from the mean.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, the use of smooth functions in GAMs is more advantageous as it allows more flexibility because it does not restrict the model to a linear relationship, although a GAM can estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for \(f(time_t\mid \beta_j)\) that allow for non-linear responses are polynomials (which can also be used in LMEMs), but a major limitation is that polynomials create a ``global'' fit as they assume that the same relationship exists everywhere, which can cause problems with inference.\textsuperscript{\protect\hyperlink{ref-beck1998}{36}} In particular, polynomial fits are known to show boundary effects because as \(t\) goes to \(\pm \infty\), \(f(time_t \mid \beta_j)\) goes to \(\pm \infty\) which is almost always unrealistic and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using a \emph{basis function} expansion of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(time_t)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the linear combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(time_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

@@

Here, the main idea is that the resulting smooth matches the data and approximates the true function without becoming too ``wiggly'' due to the noise present. A detailed exploration of wiggliness and smooth functions is beyond the scope of this manuscript, but in essence the wiggliness (or ``roughness'') of the fit is achieved by using a \emph{smoothness parameter} (\(\lambda\)), which is used to penalize the likelihood with a multiple \(\lambda\) times the integrated square of the second derivative of the spline. The second derivative of the spline is a measure of curvature, or the rate of change of the slope.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Increasing the penalty by increasing \(\lambda\) results in models with less curvature. As \(\lambda\) increases, the parameter estimates are penalized (shrunk towards 0) where the penalty reduces the wiggliness of the smooth fit to prevent overfitting. A low penalty estimate will result in wiggly functions whereas a high penalty estimate provides evidence that a linear response is appropriate.

With this in mind, if five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A. The four basis functions (and the intercept) are shown. Each of the basis functions is evaluated at six different points (because there are six points on the timeline). The coefficients for each of the basis functions of Figure \ref{fig:basis-plot}A are estimated using a penalized regression with smoothness parameter \(\lambda\) estimated when fitting the model. The penalized estimates fitted for our example are shown in Figure \ref{fig:basis-plot}B.

@@

Bayes' theorem states that the probability of an event can be calculated using prior knowledge.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the prior that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a posterior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good \emph{across the function} (``frequentist'') coverage\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}.

To understand this last part, it is worth reminding that a CI provides an estimate of the region where the ``true'' or ``mean'' value of a function exists, taking into account the randomness introduced by the sampling process. Because random samples from the population are used to calculate the ``true'' value of the function, there is inherent variability in the estimation process and the CI provides a region with a nominal value (usually, 95\%) where the function is expected to lie. In an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) if we average the coverage of the interval over the entire function we get approximately the nominal coverage (95\%). In other words, we expect that about 95\% of the points that compose the true function will be covered by the across the function CI. For this to occur then, some areas of the function must have more than nominal coverage and some areas less than the nominal coverage. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in Miller,\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Wood\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}, Simpson\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} and Marra.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

### 2021-11-30 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
Updated manuscript for GLM and GLMM notation

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. Independent errors $\varepsilon_{ijt}$ represent **random variation from the sampling process ** assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug", or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

@@

A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to account for[@pinheiro2006;@west2014]. A LMEM with interaction between time and treatment for a longitudinal study  can be written as:

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
(\#eq:LMEM)
\end{equation}

>>>>

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of treatment ($treatment_j$), time ($time_t$), and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. Independent errors $\varepsilon_{ijt}$ represent random variation from the sampling process assumed to be $\stackrel{i.i.d.}\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug", or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{1}$ and $\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

@@

A LMEM is a class of statistical models that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to account for[@pinheiro2006;@west2014]. A LMEM with interaction between time and treatment for a longitudinal study  can be written as

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
(\#eq:LMEM)
\end{equation}
